---
title: Simulate Volatile Anaesthestic Uptake as in Cowles et al. 1973
output:
    rmarkdown::html_vignette:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Simulate Volatile Anaesthestic Uptake as in Cowles et al. 1973}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{simva}
bibliography: references.bib
---

```{r setup, include = FALSE}
Sys.setenv(LANGUAGE = "en")
library("simva")
```

# Introduction

The `simva` package is an `R` implementation of the algorithm for a three
compartment model to simulate volatile anaesthetic uptake proposed in
@Cowles1973.

Here we reproduce the calulations presented in their article
for *diethyl ether*.

# Definition of the compartments

First we have to define the *partition coefficients* for the given anaesthetic
(*diethyl ether*) for all compartments.


```{r part-coefs}
part_coefs <- partition_coefficients("diethyl-ether")
```

```{r part-coefs-tbl, echo = FALSE, results = "asis"}
knitr::kable(t(part_coefs))
```

We have to define the caridac output and the proportion of cardiac output each
compartment recieves.

```{r co}
blood_flow <- cardiac_output(total = 6.3)
```

```{r co-tbl, echo = FALSE, results = "asis"}
knitr::kable(t(blood_flow))
```

Each compartment has a tissue and a blood volume.

```{r volumes}
tissue_volume <- c(
    lung_air = 2.68, lung_tissue = 1.0,
    vrg = 8.83, mus = 36.25, fat = 11.5 #, not_perfused = 7.02
)
blood_volume <- c(lung = 1.4, vrg = 3.2, mus = 0.63, fat = 0.18)
```

# Conductances and capacitances

The uptake of the volatile anaesthetic for a given compartment depends on the
properties of the anaesthetic, the air/blood flow and the volume of the
compartment.

```{r conductances}
conductances <- c(
    lung = conductance(
        flow = 4.0,                     # alveolar minute ventilation
        partition_coefficient = 1.0     # gas:gas partition coefficient
    ),
    vrg = conductance(blood_flow["vrg"], part_coefs["lung"]),
    mus = conductance(blood_flow["mus"], part_coefs["lung"]),
    fat = conductance(blood_flow["fat"], part_coefs["lung"])
)
```

```{r capacitances}
capacitances <- c(
    lung = lung_capacitance(
        tissue_volume["lung_air"],
        ## blood volume and tissue:gas == blood:gas in that gase part_coefs
        tissue_volume["lung_tissue"],
        tissue_coefficient = part_coefs["lung"],
        ## blood volume and blood:gas part_coefs
        blood_volume["lung"], part_coefs["lung"]
    ),
    vrg = capacitance(
        tissue_volume["vrg"], part_coefs["vrg"],
        blood_volume["vrg"], part_coefs["lung"]
    ),
    mus = capacitance(
        tissue_volume["mus"], part_coefs["mus"],
        blood_volume["mus"], part_coefs["lung"]
    ),
    fat = capacitance(
        tissue_volume["fat"], part_coefs["fat"],
        blood_volume["fat"], part_coefs["lung"]
    )
)
```

# Simulations

## Normal

```{r sim-normal}
normal <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.1, total_time = 10,
    conductances = conductances, capacitances = capacitances
)

knitr::kable(head(normal))

cols <- palette.colors(ncol(normal) - 1)
matplot(
    normal[, -1],
    col = cols,
    type = "l",
    lty = 1,
    ylab = "partial pressure [%]",
    xlab = "time [min]",
    xaxt = "n"
)
axis(1, 1:10 * 10, labels = 1:10)
legend(
    "topleft",
    title = "compartments",
    legend = colnames(normal)[-1],
    lty = 1,
    bty = "n",
    col = cols
)
```

## Doubled alveolar minute ventilation

```{r sim-dav}
dav <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.1, total_time = 10,
    conductances = conductances * c(2, 1, 1, 1),
    capacitances = capacitances
)
```

## Halved cardiac output

```{r sim-hco}
hco <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.1, total_time = 10,
    conductances = conductances * c(1, 0.5, 0.5, 0.5),
    capacitances = capacitances
)
```

## Humidification

```{r sim-hum}
hum <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.1, total_time = 10,
    conductances = conductances,
    capacitances = capacitances,
    use_humidification = TRUE
)
```

## Concentration effect

```{r sim-con}
con <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.1, total_time = 10,
    conductances = conductances,
    capacitances = capacitances,
    use_concentration_effect = TRUE
)
```

## Pulmonary shunt

```{r sim-psh}
psh <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.1, total_time = 10,
    conductances = conductances,
    capacitances = capacitances,
    shunt_frac = 0.1
)
```

## Metabolism

```{r sim-met}
met <- sim_anaesthetic_uptake(
    pinsp = 12, delta_time = 0.01, total_time = 10,
    conductances = conductances,
    capacitances = capacitances,
    metabolism_frac = 0.05
)
```

## Summary

```{r sim-tbl, echo = FALSE, results = "asis"}
tbl <- rbind(
    normal[100, ], dav[100, ], hco[100, ],
    hum[100, ], con[100, ], psh[100, ], met[100,]
)
rownames(tbl) <- c(
    "Normal",
    "Increased ventilation",
    "Decreased cardiac output",
    "Humidification",
    "Concentration effect",
    "Pulmonary shunt",
    "Metabolism"
)
knitr::kable(
    tbl,
    digits = 2,
    caption = paste0(
        "Summary of different simulation experiments. ",
        "Results are comparable to Table 4 in @Cowles1973."
    )
)
```

(There is a small difference in the numbers for the metabolism scenario.)

# References
